Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-23 ms/main.tex
formatting
revised result 3 corollaries

Cooperation and defection can coexist stably at frequencies $\tilde{p}^*$ and $1-\tilde{p}^*$ (\autoref{eq:vert_hori_equilibrium}) if there is horizontal transmission bias for defection, $T_A<T_B$, and the cost of cooperation is intermediate, $\gamma_2 < c < \gamma_1$; see yellow area in \autoref{fig:results2}b.
When unstable coexistence occurs, phenotype $A$ will fix if its initial frequency is $p>\tilde{p}^*$, and phenotype $B$ will fix if its initial frequency is $1-p>1-\tilde{p}^*$; this occurs when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:coexistence_without_oblique} shows the mapping $\tilde p \to \tilde p'$; see blue areas in \autoref{fig:results2}a and \autoref{fig:results2}b.

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:Intermediate_association}
Cooperation will increase when rare if social association is high enough; specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

@@

\end{result}
These conditions are illustrated in \autoref{fig:result3}a and \autoref{fig:result3}b.

Much of the literature on evolution of cooperation focuses on conditions for  initially rare cooperative phenotype to invade a population of defectors.
The next corollary deals with such a condition.
\begin{corollary}[Condition for cooperation to increase when initially rare]
  If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
  \begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
  \begin{aligned}
  c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
  \end{aligned}
  \end{equation} 
  \end{corollary}

Note that the right-hand side equals $\hat\gamma_1$.
This condition follows from combining the conditions for fixation of cooperation and for stable coexistence.

In general, these conditions cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessity condition for fixation of cooperation.]
  Fixation of cooperation is possible only when
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
When there is horizontal transmission bias for cooperation, fixation can occured for every value of vertical transmission because $\hat{v}<0$.
However, if the the horizontal bias favours defection, for fixation of cooperation to occur the vertical transmission must be higer than $\hat{v}$.
\\

We already examine the effect of social association in the case of perfect vertical transmission ($v=1$). Now we'll move to the genral case where $0<v\leq 1$.
We denote the social association boundaries of the general case by
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A}.
\end{aligned}\end{equation}
\emph{Corollary~\autoref{corollary:Intermediate_association}} is true for the general but with new boundaries coefficients $\hat{a}_1$, $\hat{a}_2$. \autoref{fig:result3}c and \autoref{fig:result3}d illustrate this result.

%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$

>>>>

Cooperation and defection can coexist stably at frequencies $\tilde{p}^*$ and $1-\tilde{p}^*$ (\autoref{eq:vert_hori_equilibrium}) if there is horizontal transmission bias for defection, $T_A<T_B$, and the cost of cooperation is intermediate, $\gamma_2 < c < \gamma_1$; see yellow area in \autoref{fig:results2}b.
When unstable coexistence occurs, phenotype $A$ will fix if its initial frequency is $p>\tilde{p}^*$, and phenotype $B$ will fix if its initial frequency is $1-p>1-\tilde{p}^*$; this occurs when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:coexistence_without_oblique} shows the mapping $\tilde p \to \tilde p'$; see blue areas in \autoref{fig:results2}a and \autoref{fig:results2}b.
\\

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:Intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

@@

\end{result}
These conditions are illustrated in \autoref{fig:result3}ab.

Much of the literature on evolution of cooperation focuses on conditions for  initially rare cooperative phenotype to invade a population of defectors.
The next corollary deals with such a condition.
\\

\begin{corollary}[Condition for cooperation to increase from rarity]
  If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
  \begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
  \begin{aligned}
  c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
  \end{aligned}
  \end{equation} 
  \end{corollary}

%Note that the right-hand side equals $\hat\gamma_1$.
This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases maximum cost of cooperation $\hat\gamma_1$ that allows cooperation to increase from rarity (i.e. $\partial \hat\gamma_1 / \partial \alpha > 0$).

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

In corollary~\autoref{corollary:Intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Then the following applies.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:Intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Interestingly, when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Therefore, when there is horizontal transmission bias for cooperation, $T_A>T_B$, then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$). In contrast, when there is horizontal transmission bias for defection, vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$).
